---
title: "N741 Homework 8 Due 04/24/2019"
author: "Melinda Higgins"
date: "April 17, 2019"
output: html_document
---

## Setup code

For this homework you will be working with the Abalone dataset from UCI, see [http://archive.ics.uci.edu/ml/datasets/abalone](http://archive.ics.uci.edu/ml/datasets/abalone). 

There were 3 dimensional measurements made of the abalones: length, diameter and height in mm (millimeters). Length was defined to be the longest length, longer than diameter or height. All measurements should be > 0.

There were also 3 weight component measurements in grams: shucked weight, viscera weight and shell weight, which are all less than the total weight of the abalones. The total whole weight is also provided.

The "sexes" of the abalones in this dataset are:

* "I" for infants (immature)
* "M" for males (adult)
* "F" for females (adult)

We will use unsupervised clustering methods in this homework to see if any of the clusters and components reflect the abalones by sex.

The dataset also includes the number of rings in the shells which, like trees, can be used to determine the age of the abalones where age = rings + 1.5. We will not be using this measure in this homework.

The setup code below:

* reads in the data, 
* adds the variable names, 
* cleans up the dataset removing cases with typos or illogical measures that do not meet the criteria specified above; and 
* the code also does a small random sample of 5% of the cases, giving us about 200+ abalones to work with for this homework exercise.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readr)

# read-in abalone dataset
# abalone <- read_csv("abalone.data", col_names = FALSE)
abalone <- read_csv(url("http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data"))

# assign variable names
names(abalone) <- c("sex","length","diameter","height",
                    "wholeWt","shuckedWt","visceraWt","shellWt",
                    "rings")

# cleanup data
# component weights should be less than whole weight
# length is defined to be the longest dimensional measurement
# and all dimensions should be > 0 - there are some 0 heights
abaloneClean <- abalone %>%
  filter(shuckedWt < wholeWt) %>%
  filter(visceraWt < wholeWt) %>%
  filter(diameter < length) %>%
  filter(height < length) %>%
  filter(height > 0)

# take a 5% random sample of 4169 cases
abaloneSample <- abaloneClean %>%
  sample_frac(size = 0.05, replace = FALSE)

# keep 3 dimensions and 3 component weights
# remove sex and number of rings
# number of rings + 1.5 = age of abalone
abaloneKeep <- abaloneSample %>%
  select(length, diameter, height, 
         shuckedWt, visceraWt, shellWt)
```

You will use the `abaloneKeep` dataset for this homework exercise, which has about 200+ cases and only has 6 variables: 3 dimensional measures and 3 component weights. 

Use the code from the Unsupervised Learning lecture to guide you in this homework - see  [https://htmlpreview.github.io/?https://github.com/vhertzb/more-supervised-learning/blob/master/More_Supervised_Learning.html](https://htmlpreview.github.io/?https://github.com/vhertzb/more-supervised-learning/blob/master/More_Supervised_Learning.html) - look about half-way down the webpage where Unsupervised Learning begins.

**NOTE: Code is provided below in place to help. REMEMBER to add `{r}` to the code chunks below so they will run as R code chunks in your final homework assignment. The `{r}` was removed to generate the homework assignment document.**

## Cluster Analysis

``` {r}
# Scale the data before clustering
sd.data <- scale(abaloneKeep)
```

PROBLEM 1: Using the scaled data `sd.data` calculate the Euclidean distance between each pair of points

```{r}
# insert code here
data.dist <- dist(sd.data)
```

PROBLEM 2: Plot the tree, default linkage = 'complete'. Set `labels = abaloneSample$sex` to get the sex labels for the plot.

```{r}
# insert code here
plot(hclust(data.dist), labels = abaloneSample$sex, main = "Complete Linkage", xlab = "", sub = "", ylab = "")
```

PROBLEM 3: Plot the tree, using linkage = 'average'. Set `labels = abaloneSample$sex` to get the sex labels for the plot.

```{r}
# insert code here
plot(hclust(data.dist), method = "average", labels = abaloneSample$sex, main = "Average Linkage", xlab = "", sub = "", ylab = "")

```

PROBLEM 4: Plot the tree, using default linkage = 'single'. Set `labels = abaloneSample$sex` to get the sex labels for the plot.

```{r}
# insert code here
plot(hclust(data.dist), method = "single", labels = abaloneSample$sex, main = "Single Linkage", xlab = "", sub = "", ylab = "")
```

PROBLEM 5: Let's use complete linkage and cut into 3 clusters. When running the `table()` line of code use `abaloneSample$sex` to compare the original sexes to the clusters `hc.clusters`.

```{r}
# insert code here

hc.out <- hclust(dist(sd.data))
hc.clusters <- cutree(hc.out, 3)
table(hc.clusters, abaloneSample$sex)
```

PROBLEM 6: Given the results above:

* which cluster(s) did the best job identifying the infants?

Cluseres 1 and 2 did the best job of identifying the infants. 

* which cluster(s) found the adults (M and F)?

Cluster 1 did the mest job of identifying both female and male. 

## K-Means clustering

PROBLEM 7: Run a K-means clustering with K=3 (same as we did above for the hierarchical clustering). Compare the clusters from K-means clustering and the hierarchical clustering `hc.clusters` (using the complete linkage method). How do these clusters compare - similar or not?

```{r}
# insert code here
set.seed(40523)
km.out = kmeans(sd.data, 3, nstart = 20)
km.clusters = km.out$cluster
table(km.clusters, hc.clusters)
```

Clusters are not similar. 

## PCA - Principal Components Analysis

PROBLEM 8: Find the principal components of the normalized data, which are computed using scale = TRUE option. Use `abaloneKeep` as the input dataset.

```{r}
# insert code here
pr.out <- prcomp(abaloneKeep, scale = TRUE)
```

### Plot of the data points on the first 2 PCs

The code to show the "Scores Plot" on the 1st two PCs (principal components) is provided below.

```{r}
# colors
cols=rainbow(length(unique(abaloneSample$sex)))

# mapping of colors to sex
# blue is #0000FFFF
# light green is #00FF00FF
# red is #FF0000FF
#
# see https://webtools.fineaty.com/ColorInfo/0000FFFF/en/Information-about-the-0000FFFF-color.html
# see https://webtools.fineaty.com/ColorInfo/00FF00FF/en/Information-about-the-00FF00FF-color.html
# see https://webtools.fineaty.com/ColorInfo/FF0000FF/en/Information-about-the-FF0000FF-color.html

table((unique(abaloneSample$sex)), cols)

# plot scores
plot(pr.out$x[, 1:2],
     col = cols)
legend(x = "topleft",
       legend = c("F", "I", "M"),
       col = cols, pch = 1)
```

PROBLEM 9: For the "Scores Plot" above, do you see any separation of the 3 sexes by their principal component scores?

_I do not see any separation of the sexes in the score plot above_

PROBLEM 10: Look at the summary of the PCA, run the summary of `pr.out`. What percentage of variance is explained by the first 2 principal components?  

```{r}
# insert code here
summary(pr.out)
```

_The cumulative proportion of the variance explained by PC1 and PC2 is 93.7%. PC1 explaings 89.7% of the variances and PC2 explains 4.0% of the variance_
PROBLEM 11: Plot the "Scree" plot

```{r}
# insert code here
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
```

## MDS - Multidimensional Scaling

PROBLEM 12: Compute the Euclidean distance between each pair of points for `abaloneKeep` (you do not need to scale the data first - just use `abaloneKeep` as is).

```{r}
# insert code here
data.dist2 <- dist(abaloneKeep)
```

PROBLEM 13: Run `cmdscale()` function to perform a Classical Metric MDS, set `eig-TRUE` and `k=2`. Save the results in an object named `fit`. View the results for `fit`.

```{r}
# insert code here
fit <- cmdscale(data.dist2, eig=TRUE, k=2)
fit
```

PROBLEM 14: Plot the MDS fit results, set `labels = abaloneSample$sex`.

```{r}
# insert code here
x <- fit$points[,1]
y <- fit$points[,2]

plot(x, y, xlab = "PCo1", ylab = "PCo2",
     main = "Metric MDS", type = "n")
text(x, y, labels = abaloneSample$sex, cex = 0.7)
```



